---
title: "A Musical Introduction to Generators"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{A Musical Introduction to Generators}
  %\VignetteEncoding{UTF-8}
---

# A Musical Introduction to Generators

What a generator allows you to do is to take code that is writen in
a sequential, looping style, and then treat that code's output as
a collection to be iterated over.

To illustrate what this means, and the the generators package in
general, I will use R to perform a suitable piece of music,
specifically Steve Reich's "Clapping Music."

The piece starts with a loop counting out groups of 3, 2, 1, 2, 3, 2,
1, 2, 3... with each group separated by one rest. This adds up to a
12-note loop:

![Musical score showing loop](clapping_music.svg)

In code, we could implement that idea like the following, printing a
`1` for each clap and a `0` for each rest:

```{R}
print_pattern <- function(counts = c(3, 2, 1, 2), repeats=4) {
  for (i in seq_len(repeats)) {
    for (c in counts) {
        for (j in 1:c)
          cat(1)
        cat(0)
    }
  }
  cat("\n")
}
```

Testing this, we should see groups of 1, 2, or 3 `1`s separated by a single `0`:
```{R}
print_pattern(repeats=4);
```

"Clapping Music" is based on manipulating this 12-count loop. But it's hard to
manipulate the output of a program that only prints. The calls
to `cat` produce output on the terminal, but they don't produce data
that we can easily manipulate with more programming -- we need to make
this pattern into data, rather than terminal output.

The `generators` package allows us to enclose a data-generating process into an
object.

To make a generator for this pattern, we just enclose the body of the
function in a call to `gen()`, and change each `cat()` to
`yield()`. We can also move the outer loop into the `recycle` argument
of an [iteror].

```{R}
library(async)
counts <- c(3, 2, 1, 2)
pattern <- gen({
  repeat {
    for (n in counts) {
      for (j in 1:n)
        yield(1)
      yield(0)
    }
  }
})
```

The code inside `gen(...)` does not run, yet. The call to `gen`
constructs an [iteror][]. Iteror is an implementation of iterators
defined in the async package; its objects are cross compatible with
the established `iterators` package but can have somewhat better
performance. The main method for an iteror is `nextElemOr(x, or)`
where the second argument specifies what to do to signal the end
of iteration. So we can get out our pattern like this:

```{R}
for (i in 1:24) {
  cat(nextElemOr(pattern, break))
}
cat("\n")
```

When `nextElemOr()` is called on a generator, the generator runs its
code only up to where `yield` is called. The generator returns this
value, and saves its state, pausing until the next call to `nextElemOr()`.

We'll have to create a few copies of this kind of generator, so we can reuse it
by making a function. You can wrap `gen` around a function
expression; this will define a _generator function_ (a function that
constructs a generator.)

```{R}
gen_pattern <- gen(function(counts = c(3, 2, 1, 2)) {
  repeat {
    for (n in counts) {
      for (j in 1:n)
        yield(1)
      yield(0)
    }
  }
})
```

Because `gen(...)` builds an [iteror][], you you can apply iteror
methods to it. For instance you can collect just the first 24 items
with `ilimit()`:

```{R}
show_head <- function(x, n=24) {
  x |> ilimit(n) |> as.list() |> deparse() |> cat(sep="\n")
}
show_head(gen_pattern(), 24)
```

## Making noise

We're a good way into what I advertised as a musical endeavour and
haven't made any sounds yet. First let's download some handclap
samples. I located some on GitHub:

```{R, eval=FALSE}
tmp <- tempdir()
baseurl <- "https://github.com/octoblu/drum-kit/raw/master/public/assets/samples"
samplepaths <- paste0(tmp, c("x" = "/clap4.wav","X" = "/clap5.wav"))
curl::curl_download(paste0(baseurl, "/clap%20(4).WAV"), samplepaths[1])
curl::curl_download(paste0(baseurl, "/clap%20(5).WAV"), samplepaths[2])
```

Although R is not really known for audio performance, there is an `audio`
package playing sound samples, which we can use like this:

```{R, eval=FALSE}
library(audio) # for load.wave, play
claps <- lapply(samplepaths, load.wave)
play(claps[[1]])
play(claps[[2]])
```

We want to play sounds at a consistent tempo, so here's a routine that
takes in a generator and a sample list, and plays at a given
tempo. The `profvis` package has a `pause` function that's more
accurate than `Sys.sleep()`.

```r
library(profvis) # for pause
iplay <- function(g, samples, bpm) {
  interval <- 60 / bpm
  target <- Sys.time()
    repeat {
      x <- nextElemOr(g, break)
      target <- target + interval
      while({now <- Sys.time(); Sys.time() - now > 0.15})
        Sys.sleep(target - now - 0.15)
      if (is.numeric(x) && x >= 1 && x <= length(samples)) {
        cat(x)
        pause(target - Sys.time())
        play(samples[[x]])
      } else {
        cat(".")
      }
    }
}
```

So we should hear our pattern now:

```{R, eval=FALSE}
gen_pattern() |> ilimit(36) |> iplay(claps, 360)
```

### Some iteror (iterator) functions

The object constructed by `gen` has class `iteror`. Iterors are an
iteration construct included in this package. Iterors are cross
compatible with iterators from the `iterators` package and its
`nextElem` method, but using their preferred method `nextElemOr` can
lead to better performance and more compact code.

We can work with generators using iteror methods. Here are two methods
that will come in handy. One is an iteror equivalent of `lapply`,
which I'll call `iapply`. The other one is `isink` which just consumes
and throws away all elements from an iteror.

```{R}
iapply <- function(it, f, ...) { list(it, f, ...)
  iteror(function(or) {
    f(nextElemOr(it, return(or)), ...)
  })
}

isink <- function(it, then=invisible(NULL)) {
  repeat nextElemOr(it, break)
  then
}
```

Note that the second argument to `nextElemOr` is only lazily evaluated
when the iterator terminates; this is what allows you to place control
flow operators like `break`.

For example, we can print the contents of a generator by `iapply`ing
`cat` and throwing away the results:

```{R}
g <- gen_pattern() |> ilimit(24) |> iapply(cat) |> isink(cat("\n"))
```


### Some iterator functions, using generators

Generators enjoy some syntax extensions over base R; one extension is
that they can use a regular `for` loop with an iteror. So we can
equivalently write the above `iapply` and `isink` using even more
concise generator syntax.

```{R}
gapply <- gen(function(it, f, ...) for (x in it) yield(f(x, ...)) )
gsink <- function(it, then=NULL) {run(for (i in it) NULL); then}
```

Here [run][] is like `gen`, except rather than constructing an iteror,
`run` executes immediately, collecting all values passed to `yield`
and returning a list. So `gsink` is a function that when called will
consume all items from the iterator `it` before returning.

## Phasing and combining

"Clapping Music" is a piece for two performers, who both play the same
pattern, but after every 12 loops, one of the performers skips forward
by one step. Over the course of the piece, the two parts move out and
back into in phase with each other. We can write a generator function
that does this "skip," by consuming a value without yielding it:

```{R}
drop_one_after <- gen(function(g, n, sep=character(0)) {
  repeat {
    for (i in 1:n) yield(nextElemOr(g, break))
    nextElemOr(g, break) #drop
    cat(sep) # print a seperator after every skip
  }
})
```

Here's what 12 loops look like, where you after three (i.e. skipping every fourth):

```{R}
iseq() |>
ilimit(12) |>
drop_one_after(3, "\n") |>
iapply(cat, "") |>
isink()
```

The performance directions for "Clapping Music" request that the two
performers should make their claps sound similar, so that their lines
blend into an overall pattern. We can interpret that as combining the
two lines by adding two generators, resulting in 0, 1, or 2 claps at
every step, playing the louder sample for a value of 2.

Then, all together:

```{r}
clapping_music <- function(n=12, counts=c(3,2,1,2), sep=" ") {
  cell <- sum(counts+1) # how long?
  a <- gen_pattern(counts)
  b <- gen_pattern(counts) |> drop_one_after(n*cell, sep)
  # add them together and limit the output
  gen(for (i in 1:(n*(cell+1)*cell)) {
    x <- nextElemOr(a, break)
    y <- nextElemOr(b, break)
    yield(x+y)
  })
}
```

To narrate this: we are constructing two independent instances of our
12-note generator. One of these patterns is made to skip one beat every N
bars. Then we create a third generator that adds together the two.

```{R}
clapping_music(4, sep="\n") |> iapply(cat) |> isink(cat("\n"))
```

## A performance

Now we should be able to hear our performance:

```{r, eval=FALSE}
iplay(clapping_music(n=4, sep="\n"), claps, 480)
```

It has to be said that R is not particularly built to be a multimedia
environment, plus the `audio` package will be using the OS alert sound
facility, which is typically not built for precise timing, so you may
hear some glitches and hiccups. Nevertheless, I hope this has
illustrated how generators allows control to be _interleaved_ among
different sequential processes.
